### 2016-08-15 Chapters/Ch03-PCMCoupledCluster.tex
Not much...
Save work on CC chapter

% mainfile: ../RobertoDiRemigioPhDThesis.tex
\chapter{Coupled cluster}\label{ch:PCMCC}

@@

@@

   \sigma^\mathrm{CC} \\
   \kappa^\mathrm{CC}

@@

  \frac{1}{2}(\sigma^\mathrm{HF}+\sigma^\mathrm{CC})\PCM(\sigma^\mathrm{HF}+\sigma^\mathrm{CC})
+ (\sigma^\mathrm{HF}+\sigma^\mathrm{CC})(\esp^\mathrm{HF}+\tilde{\esp}(\tamp{},\tbar{})) \\
&+ \frac{1}{2}(\kappa^\mathrm{HF}+\kappa^\mathrm{CC})\MM(\kappa^\mathrm{HF}+\kappa^\mathrm{CC})
+ (\kappa^\mathrm{HF}+\kappa^\mathrm{CC})(\zeta^\mathrm{HF}+\tilde{\zeta}(\tamp{},\tbar{})) \\
&+ (\sigma^\mathrm{HF}+\sigma^\mathrm{CC})\bi{X}(\kappa^\mathrm{HF}+\kappa^\mathrm{CC}) \\

@@

&+  \frac{1}{2}\sigma^\mathrm{CC}\PCM\sigma^\mathrm{CC}
  + \sigma^\mathrm{CC}\tilde{\esp}(\tamp{},\tbar{})
  + \frac{1}{2}\kappa^\mathrm{CC}\MM\kappa^\mathrm{CC}
  + \kappa^\mathrm{CC}\tilde{\zeta}(\tamp{},\tbar{})
  + \sigma^\mathrm{CC}\bi{X}\kappa^\mathrm{CC} \\
&+  \frac{1}{2}\sigma^\mathrm{HF}\PCM\sigma^\mathrm{CC}

@@

  + \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa^\mathrm{CC}

@@

  + \sigma^\mathrm{HF}\bi{X}\kappa^\mathrm{CC} \\
&+  \frac{1}{2}\sigma^\mathrm{CC}\PCM\sigma^\mathrm{HF}
  + \sigma^\mathrm{CC}\esp^\mathrm{HF}
  + \frac{1}{2}\kappa^\mathrm{CC}\MM\kappa^\mathrm{HF}
  + \kappa^\mathrm{CC}\zeta^\mathrm{HF}
  + \sigma^\mathrm{CC}\bi{X}\kappa^\mathrm{HF} \\

@@

&+  \frac{1}{2}\sigma^\mathrm{CC}\PCM\sigma^\mathrm{CC}
  + \sigma^\mathrm{CC}\tilde{\esp}(\tamp{},\tbar{})
  + \frac{1}{2}\kappa^\mathrm{CC}\MM\kappa^\mathrm{CC}
  + \kappa^\mathrm{CC}\tilde{\zeta}(\tamp{},\tbar{})
  + \sigma^\mathrm{CC}\bi{X}\kappa^\mathrm{CC} \\
&+ \sigma^\mathrm{CC}(\bi{Y}\sigma^\mathrm{HF} + \bi{X}\kappa^\mathrm{HF} + \esp^\mathrm{HF}) \\
&+ \kappa^\mathrm{CC}(\bi{W}\kappa^\mathrm{HF} + \bi{X}^\dagger\sigma^\mathrm{HF} + \zeta^\mathrm{HF}) \\

@@

&+  \frac{1}{2}\sigma^\mathrm{CC}\PCM\sigma^\mathrm{CC}
  + \sigma^\mathrm{CC}\tilde{\esp}(\tamp{},\tbar{})
  + \frac{1}{2}\kappa^\mathrm{CC}\MM\kappa^\mathrm{CC}
  + \kappa^\mathrm{CC}\tilde{\zeta}(\tamp{},\tbar{})
  + \sigma^\mathrm{CC}\bi{X}\kappa^\mathrm{CC} \\

@@

where we made use of the stationarity condition for the reference
reaction field to eliminate a number of terms. Notice that the
reference reaction energy still appears among the terms in the
functional. However, these terms will not enter in the optimization of
the \acs{CC} wave function and we will therefore drop them in the
following, out of notational convenience:

@@

  \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{}) \\
  &= \frac{1}{2}\p^\dagger\V\p + \p^\dagger\tilde{\s}(\tamp{},\tbar{})
  + \p^\mathrm{HF}\tilde{\s}(\tamp{},\tbar{})

@@

One can then write the effective Lagrangian as:
\begin{equation}
  \begin{aligned}
  \lag{M}(\tamp{}, \tbar{}, \sigma) &=
  \braket{\mathrm{HF} | \overline{H_0} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tamp{i} | \overline{H_0} | \mathrm{HF}} \\
  &+ \frac{1}{2}\sigma\PCM\sigma
  + \sigma\tilde{\esp}(\tamp{}, \tbar{})
  + \frac{1}{2}\kappa\MM\kappa
  + \kappa\tilde{\zeta}(\tamp{}, \tbar{})
  + \sigma\bi{X}\kappa
  + \sigma^\mathrm{HF}\tilde{\esp}(\tamp{}, \tbar{})
  + \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{}, \tbar{})
  \end{aligned}
\end{equation}
In accordance with previous work,\autocite{Cammi2009-gu, Caricato2011-tx} the
Hamiltonian $H_0$ is augmented with the terms
$\sigma^\mathrm{HF}\tilde{\esp}(\tamp{}, \tbar{})$ and
$\kappa^\mathrm{HF}\tilde{\zeta}(\tamp{}, \tbar{})$
part appearing in the previous functional yielding the \acl{PTE} (\acs{PTE}) Hamiltonian --
$H$ -- and
eventually the \acl{PTED} (\acs{PTED}) Lagrangian:
\begin{equation}\label{eq:pted-cc}
  \begin{aligned}
  \lag{M}(\tamp{}, \tbar{}, \sigma) &=
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{H} | \mathrm{HF}} \\
  &+ \frac{1}{2}\sigma\PCM\sigma
  + \sigma\tilde{\esp}(\tamp{}, \tbar{})
  + \frac{1}{2}\kappa\MM\kappa
  + \kappa\tilde{\zeta}(\tamp{}, \tbar{})
  + \sigma\bi{X}\kappa \\
  &=
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{H} | \mathrm{HF}}
  + \frac{1}{2}\p^\dagger\V\p + \p^\dagger\tilde{\s}(\tamp{},\tbar{})
  \end{aligned}
\end{equation}
As noted by \citeauthor{Cammi2009-gu} for the implicit \acs{PCM} model,
the \acs{PTE} Hamiltonian is computed in \acs{MP}-partitioned form by
employing the "solvated" orbitals and Fock matrix:
\begin{equation}
  H = F + \Phi.
\end{equation}
$F$ is a diagonal one-electron operator and $\Phi$ is the
fluctuation potential.
The \acs{PTE} effective Lagrangian ${}^\mathrm{PTE}\lag{M}$ can be
recovered by dropping the classical terms in the Lagrangian
\eqref{eq:pted-cc}:
\begin{equation}
{}^\mathrm{PTE}\lag{M}
  =
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{H} | \mathrm{HF}}
\end{equation}
in accordance with the derivation of \citeauthor{Cammi2009-gu}.
As shown by \citeauthor{Caricato2011-tx}, it is possible to achieve less
drastic approximations of the \acs{PTED} Lagrangian. Such approximations
can be easily derived within the current variational framework. We
postpone this analysis to a later stage of development, \emph{vide
infra}.

@@

In the supermatrix formalism, the similarity-transformed terms will be
written as:
\begin{equation}
  \overline{\s} = \s + \tilde{\s} = \s +
  \commutator{\s}{T} +
  \frac{1}{2}\commutator{\commutator{\s}{T}}{T}
\end{equation}
It is important to single out which terms contribute in the various
expectation values where the similarity-transformed sources will be
involved.
We develop this analysis in terms of excitation ranks of the
operators involved, based on the following:
\begin{equation}
  \braket{\mu_n |
  \commutator{\commutator{\commutator{O}{T_{n_1}}}{\ldots}}{T_{n_k}} | \mathrm{HF}} \neq 0
  \Longleftrightarrow
  n - m_O + k \leq \sum_{i=1}^k n_i \leq n + m_O
\end{equation}
where $n$ is the excitation rank of the $\ket{\mu_n}$ determinant, $k$
is the total number of cluster operators appearing in the nested
commutator, the $n_i$ are their respective excitations ranks and $m_O$
is the \emph{particle rank} of the operator $O$.\autocite{Helgaker2000-tz}
This follows from the realization that:
\begin{equation}
  \commutator{\commutator{\commutator{O}{T_{n_1}}}{\ldots}}{T_{n_k}} \ket{\mathrm{HF}}
\end{equation}
is a linear combination of determinants with excitation ranks $v$ in the
range:
\begin{equation}
  \sum_{i=1}^k n_i - m_O \leq v \leq \sum_{i=1}^k n_i + m_O -k
\end{equation}
The source operator is a nondiagonal, one-electron
operator and hence has particle rank $m_\s = 1$.

Eventually, one can cast the source terms entering the Lagrangian as
follows:
\begin{equation}
  \tilde{\s}(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \s + \BCHfirst{\s}{T} + \BCHsecond{\esp}{T} | \mathrm{HF}}
\end{equation}

\begin{align}
  \braket{\mathrm{HF} |
  \BCHfirst{\s}{T_t}
  | \mathrm{HF}}
  &\quad\quad\text{nonzero contributions} \quad 0 \leq t \leq 1
  \Longleftrightarrow t = 1 \\
  \braket{\mathrm{HF} |
  \commutator{\commutator{\s}{T_t}}{T_u}
  | \mathrm{HF}}
  &\quad\quad\text{nonzero contributions} \quad t + u = 1 \\
  \braket{\tbar{i} |
  \esp
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad i = 1 \\
  \braket{\tbar{i} |
  \BCHfirst{\s}{T_t}
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad i \leq t \leq i + 1 \\
  \braket{\tbar{i} |
  \commutator{\commutator{\s}{T_t}}{T_u}
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad t + u = i + 1
\end{align}

For a general truncation level we accordingly obtain:
\begin{equation}\label{eq:source-term}
  \tilde{\s}(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
\end{equation}
which shows the unique role the singles amplitudes assume when a
semiclassical environment model is introduced, \emph{vide infra}.
Furthermore, for the \acs{CCS} model ($\mathcal{M} = 1$):
\begin{equation}
  \tilde{\s}^1(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}}
  + \braket{\tbar{1} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \frac{1}{2}\braket{\tbar{1} | \commutator{\commutator{\s}{T_1}}{T_1} | \mathrm{HF}}
\end{equation}
while for the \acs{CCSD} model ($\mathcal{M} = 2$):
\begin{equation}
  \tilde{\s}^2(\tamp{}, \tbar{}) =
  \tilde{\s}^1(\tamp{}, \tbar{})
  + \braket{\tbar{1} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} | \mathrm{HF}}
\end{equation}
Eventually, within the \acs{CCSDT} model ($\mathcal{M} = 3$) four more
terms are added:
\begin{equation}
  \begin{aligned}
  \tilde{\s}^3(\tamp{}, \tbar{}) &=
  \tilde{\s}^2(\tamp{}, \tbar{})
  + \braket{\tbar{2} | \BCHfirst{\s}{T_3} | \mathrm{HF}} \\
  &+ \braket{\tbar{3} | \BCHfirst{\s}{T_3} | \mathrm{HF}}
  + \frac{1}{2}
  \braket{\tbar{3} | \commutator{\commutator{\s}{T_2}}{T_2} |
  \mathrm{HF}}
  + \braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} |
  \mathrm{HF}}
  \end{aligned}
\end{equation}
We note that:
\begin{itemize}
    \item double commutators involving the highest order cluster
      operators can only lead to nonzero terms when $T_1$ is also
      involved and by projection onto the highest order left \acs{CC}
      state, \emph{e.g.}
      $\braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} | \mathrm{HF}}$
      and
      $\braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} | \mathrm{HF}}$
\end{itemize}

\section{The PTED-CC equations}\label{sec:pted-cc-equations}

>>>>

% mainfile: ../RobertoDiRemigioPhDThesis.tex
%************************************************
\chapter{Treating solvation within correlated wave function models}\label{ch:PCMCC}

\epigraph{Non temerai i terrori della notte non temerai il terrore}{--- \textsc{CCCP}}

In this Chapter, I will summarize previous and current work on the
formulation of quantum/classical polarizable models in the context of
the highly accurate \acl{CC} model.
This is an ongoing and yet unpublished collaboration with prof.~T.~Daniel Crawford
(Virginia Tech) and Dr.~Andrew C.~Simmonett (NIH)
The first part of this Chapter summarizes existing work, mainly by Cammi
and Caricato, by recasting it in language of variational
quantum/classical polarizable Hamiltonians.
The second part of this Chapter illustrates how the variational language
facilitates the formulation of a \acl{CC}-based \acl{PT} (\acs{CC}-\acs{PT})
including a classical polarizable environment.
I will finally show how the perturbation series can be successfully
exploited to derive iterative and noniterative approximate \acs{CC}
schemes.

\pagebreak

@@

   \sigma^\mathrm{tot} \\
   \kappa^\mathrm{tot}

@@

@@

  \frac{1}{2}(\sigma^\mathrm{HF}+\sigma)\PCM(\sigma^\mathrm{HF}+\sigma)
+ (\sigma^\mathrm{HF}+\sigma)(\esp^\mathrm{HF}+\tilde{\esp}(\tamp{},\tbar{})) \\
&+ \frac{1}{2}(\kappa^\mathrm{HF}+\kappa)\MM(\kappa^\mathrm{HF}+\kappa)
+ (\kappa^\mathrm{HF}+\kappa)(\zeta^\mathrm{HF}+\tilde{\zeta}(\tamp{},\tbar{})) \\
&+ (\sigma^\mathrm{HF}+\sigma)\bi{X}(\kappa^\mathrm{HF}+\kappa)
  %%%% Steps
= \left/ \text{Expand and collect} \right/ \\

@@

&+  \frac{1}{2}\sigma\PCM\sigma
  + \sigma\bi{X}\kappa \\
&+  \frac{1}{2}\sigma^\mathrm{HF}\PCM\sigma

@@

  + \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa

@@

  + \sigma^\mathrm{HF}\bi{X}\kappa \\
&+  \frac{1}{2}\sigma\PCM\sigma^\mathrm{HF}
  + \sigma\esp^\mathrm{HF}
  + \frac{1}{2}\kappa\MM\kappa^\mathrm{HF}
  + \kappa\zeta^\mathrm{HF}
  + \sigma\bi{X}\kappa^\mathrm{HF} \\
  %%%% Steps
&= \left/
  \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa^\mathrm{HF}
  \sigma^\mathrm{HF}\esp^\mathrm{HF}
  \kappa^\mathrm{HF}\zeta^\mathrm{HF}
  \sigma^\mathrm{HF}\bi{X}\kappa^\mathrm{HF}
  \frac{1}{2}\kappa^\mathrm{HF}\zeta^\mathrm{HF}
  \right/ \\

@@

&+  \frac{1}{2}\sigma\PCM\sigma
  + \sigma\bi{X}\kappa \\
&+  \frac{1}{2}\sigma^\mathrm{HF}\PCM\sigma
  + \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa
  + \sigma^\mathrm{HF}\bi{X}\kappa \\
&+  \frac{1}{2}\sigma\PCM\sigma^\mathrm{HF}
  + \sigma\esp^\mathrm{HF}
  + \frac{1}{2}\kappa\MM\kappa^\mathrm{HF}
  + \kappa\zeta^\mathrm{HF}
  + \sigma\bi{X}\kappa^\mathrm{HF} \\
  %%%% Steps
&= \left/
  \frac{1}{2}\sigma\PCM\sigma^\mathrm{HF}
  \frac{1}{2}\sigma^\mathrm{HF}\PCM\sigma;\quad
  \frac{1}{2}\kappa\MM\kappa^\mathrm{HF}
  \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa; \quad
  \sigma^\mathrm{HF}\bi{X}\kappa
  \kappa\bi{X}^\dagger\sigma^\mathrm{HF}
  \right/ \\

@@

&+  \frac{1}{2}\sigma\PCM\sigma
  + \sigma\bi{X}\kappa \\
  &+ \sigma(\cancelto{0}{\bi{Y}\sigma^\mathrm{HF} + \bi{X}\kappa^\mathrm{HF} + \esp^\mathrm{HF}})
  + \kappa(\cancelto{0}{\bi{W}\kappa^\mathrm{HF} + \bi{X}^\dagger\sigma^\mathrm{HF} + \zeta^\mathrm{HF}}) \\
  %%%% Steps
&= \left/
  + \frac{1}{2}\kappa^\mathrm{HF}\zeta^\mathrm{HF}
  \coloneqq U_\mathrm{pol}^\mathrm{ref}
  \right/ \\
  U_\mathrm{pol}^\mathrm{ref}
  +  \frac{1}{2}\sigma\PCM\sigma
  + \sigma\bi{X}\kappa \\

@@

Notice that the reference reaction energy still appears among the terms
in the functional. Obviously, these terms will not enter in the optimization of
the \acs{CC} wave function.
Introducing the supermatrix formalism, further simplifies the
polarization functional:

@@

  \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{})
  + U_\mathrm{pol}^\mathrm{ref}
  \frac{1}{2}\p^\dagger\V\p + \p^\dagger\tilde{\s}(\tamp{},\tbar{})
  + \p^{\mathrm{HF}, \dagger}\tilde{\s}(\tamp{},\tbar{})
  + U_\mathrm{pol}^\mathrm{ref}

@@

One can then write the effective Lagrangian as:
\begin{equation}
  \begin{aligned}
  \lag{M}(\tamp{}, \tbar{}, \p) &=
  \braket{\mathrm{HF} | \overline{H_0} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tamp{i} | \overline{H_0} | \mathrm{HF}} \\
  &+
  \frac{1}{2}\p^\dagger\V\p + \p^\dagger\tilde{\s}(\tamp{},\tbar{})
  + \p^{\mathrm{HF}, \dagger}\tilde{\s}(\tamp{},\tbar{})
  + U_\mathrm{pol}^\mathrm{ref}
  \end{aligned}
\end{equation}
In accordance with previous work,\autocite{Cammi2009-gu, Caricato2011-tx} the
Hamiltonian $H_0$ is augmented with the
$\p^{\mathrm{HF},\dagger}\tilde{\s}(\tamp{},\tbar{})$ term
to yield the \acl{PTE} (\acs{PTE}) Hamiltonian -- $H$ -- and the
eventually the \acl{PTED} (\acs{PTED}) Lagrangian:
\begin{equation}\label{eq:pted-cc}
  \lag{M}(\tamp{}, \tbar{}, \p) =
  {}^\mathrm{PTE}\lag{M}
  + \frac{1}{2}\p^\dagger\V\p + \p^\dagger\tilde{\s}(\tamp{},\tbar{})
\end{equation}
where the \acs{PTE} Lagrangian has been introduced:
\begin{equation}
{}^\mathrm{PTE}\lag{M}
  =
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{H} | \mathrm{HF}}
  + U_\mathrm{pol}^\mathrm{ref}
\end{equation}

As noted by \citeauthor{Cammi2009-gu} for the implicit \acs{PCM} model,
the \acs{PTE} Hamiltonian is computed in \acs{MP}-partitioned form by
employing the ``solvated'' orbitals and Fock matrix:
\begin{equation}
  H = F + \Phi.
\end{equation}
$F$ is a diagonal one-electron operator and $\Phi$ is the
fluctuation potential.
The simple nature of the Fock operator lets us re-express the \acs{PTE}
Lagrangian as follows:
\begin{equation}
{}^\mathrm{PTE}\lag{M}
  =
  G_0
  + \sum_{i=1}^{\mathcal{M}}\tbar{\mu_i}\denom{\mu_i}\tamp{\mu_i}
  + \braket{\mathrm{HF} | \overline{\Phi} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\Phi} | \mathrm{HF}}
\end{equation}
where $G_0 = E_0 + U_\mathrm{pol}^\mathrm{ref}$ is the reference free
energy.

Use of the \acs{PTE} Lagrangian yields \acs{CC} equations that are
completely equivalent in form to those obtained \emph{in vacuo}.
The model is also easily implemented, as it only requires access to a
reference state optimized in the presence of the classical medium.
The \acs{PTE} model thus offers an efficient approximation to the full
\acs{PTED} model with a computational cost comparable to that of
\emph{in vacuo} \acs{CC} theory.
As shown by \citeauthor{Caricato2011-tx}, it is possible to achieve less
drastic approximations of the \acs{PTED} Lagrangian. Such approximations
can be easily derived within the current variational framework. We
postpone this analysis to a later stage of development, \emph{vide
infra}.

@@

In the supermatrix formalism, the similarity-transformed terms will be
written as:
\begin{equation}
  \overline{\s} = \s + \tilde{\s} = \s +
  \commutator{\s}{T} +
  \frac{1}{2}\commutator{\commutator{\s}{T}}{T}
\end{equation}
It is important to single out which terms contribute in the various
expectation values where the similarity-transformed sources will be
involved.
We develop this analysis in terms of excitation ranks of the
operators involved, based on the following:
\begin{equation}
  \braket{\mu_n |
  \commutator{\commutator{\commutator{O}{T_{n_1}}}{\ldots}}{T_{n_k}} | \mathrm{HF}} \neq 0
  \Longleftrightarrow
  n - m_O + k \leq \sum_{i=1}^k n_i \leq n + m_O
\end{equation}
where $n$ is the excitation rank of the $\ket{\mu_n}$ determinant, $k$
is the total number of cluster operators appearing in the nested
commutator, the $n_i$ are their respective excitations ranks and $m_O$
is the \emph{particle rank} of the operator $O$.\autocite{Helgaker2000-tz}
This follows from the realization that:
\begin{equation}
  \commutator{\commutator{\commutator{O}{T_{n_1}}}{\ldots}}{T_{n_k}} \ket{\mathrm{HF}}
\end{equation}
is a linear combination of determinants with excitation ranks $v$ in the
range:
\begin{equation}
  \sum_{i=1}^k n_i - m_O \leq v \leq \sum_{i=1}^k n_i + m_O -k
\end{equation}

We first expand the source terms entering the Lagrangian as follows:
\begin{equation}
  \tilde{\s}(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \s + \BCHfirst{\s}{T} +
  \BCHsecond{\esp}{T} | \mathrm{HF}},
\end{equation}
and since $\s$ has particle rank $m_\s = 1$ we can conclude that the
following relationships hold:
\begin{subequations}
\begin{align}
  \braket{\mathrm{HF} |
  \BCHfirst{\s}{T_t}
  | \mathrm{HF}}
  &\quad\quad\text{nonzero contributions} \quad 0 \leq t \leq 1
  \Longleftrightarrow t = 1 \\
  \braket{\mathrm{HF} |
  \commutator{\commutator{\s}{T_t}}{T_u}
  | \mathrm{HF}}
  &\quad\quad\text{nonzero contributions} \quad t + u = 1 \\
  \braket{\tbar{i} |
  \s
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad i = 1 \\
  \braket{\tbar{i} |
  \BCHfirst{\s}{T_t}
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad i \leq t \leq i + 1 \\
  \braket{\tbar{i} |
  \commutator{\commutator{\s}{T_t}}{T_u}
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad t + u = i + 1
\end{align}
\end{subequations}
For a general truncation level we accordingly obtain:
\begin{equation}\label{eq:source-term}
  \tilde{\s}(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
\end{equation}
Furthermore, for the \acs{CCS} model ($\mathcal{M} = 1$):
\begin{equation}
  \tilde{\s}^1(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}}
  + \braket{\tbar{1} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \frac{1}{2}\braket{\tbar{1} | \commutator{\commutator{\s}{T_1}}{T_1} | \mathrm{HF}}
\end{equation}
while for the \acs{CCSD} model ($\mathcal{M} = 2$):
\begin{equation}
  \tilde{\s}^2(\tamp{}, \tbar{}) =
  \tilde{\s}^1(\tamp{}, \tbar{})
  + \braket{\tbar{1} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} | \mathrm{HF}}
\end{equation}
Eventually, within the \acs{CCSDT} model ($\mathcal{M} = 3$) four more
terms are added:
\begin{equation}
  \begin{aligned}
  \tilde{\s}^3(\tamp{}, \tbar{}) &=
  \tilde{\s}^2(\tamp{}, \tbar{})
  + \braket{\tbar{2} | \BCHfirst{\s}{T_3} | \mathrm{HF}} \\
  &+ \braket{\tbar{3} | \BCHfirst{\s}{T_3} | \mathrm{HF}}
  + \frac{1}{2}
  \braket{\tbar{3} | \commutator{\commutator{\s}{T_2}}{T_2} |
  \mathrm{HF}}
  + \braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} |
  \mathrm{HF}}
  \end{aligned}
\end{equation}
We note that:
\begin{itemize}
  \item singles amplitudes assume a unique role when a quantum/classical
    Hamiltonian is introduced, a role that will be more evident in the
    perturbative treatment,
  \item double commutators involving the highest order cluster
    operators can only lead to nonzero terms when $T_1$ is also
    involved and by projection onto the highest order left \acs{CC}
    state, \emph{e.g.}
    $\braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} |
    \mathrm{HF}} \neq 0$
    and
    $\braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} |
    \mathrm{HF}} \neq 0$
\end{itemize}

\section{A family of PCM-CC models}\label{sec:pcm-cc-models}

Differentiation of the effective Lagrangian in Eq.~\eqref{eq:pted-cc}
with respect to the variational parameters $\tamp{\mu_i}$,
$\tbar{\mu_i}$ and $\p$ yields the \acs{PTED}-\acs{CC}
equations:
\begin{subequations}
  \begin{align}
   \tampEq{\mu_i}(\tamp{}, \tbar{}, \p)  &=
   \denom{\mu_i}\tamp{\mu_i} + \braket{\mu_i | \overline{\Phi} | \mathrm{HF}}
   + \p^\dagger\braket{\mu_i | \overline{\s} | \mathrm{HF}}
             = 0 \label{eq:pted-cc-amplitudes}\\
   \tbarEq{\mu_i}(\tamp{}, \tbar{}, \p)
    &=
    \denom{\mu_i}\tbar{\mu_i} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} |
    \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}}
    \nonumber \\
    &+ \p^\dagger\lbrace
    \braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_i}} +
    \commutator{\commutator{\s}{T}}{\cluster{\mu_i}}| \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} | \commutator{\overline{\s}}{\cluster{\mu_i}} | \mathrm{HF}}
             \rbrace
             = 0 \label{eq:pted-cc-multipliers}\\
    \Omega_\p(\tamp{}, \tbar{}, \p)
    &=
    \V\p + \tilde{\s}(\tamp{}, \tbar{}) = 0
    \label{eq:pted-cc-polarization}
  \end{align}
\end{subequations}
These equations clearly show:
\begin{enumerate}
  \item The $\tilde{\s}(\tamp{}, \tbar{})$ expectation value term in the
    polarization equation couples the amplitudes, multipliers and
    polarization equations.
    This requires a proper macroiteration/microiteration
    self-consistency scheme for its implementation.~\autocite{Cammi2009-gu, Caricato2010-hx}
  \item As a consequence, a single-point \acs{PTED}-\acs{CC} will
    suffer from a $2\times$ prefactor in its computational cost with
    respect to an \emph{in vacuo} \acs{CC} calculation.
  \item Eq.~\eqref{eq:pted-cc-polarization} is \emph{directly} coupled to
    the multipliers equation, Eq.~\eqref{eq:pted-cc-multipliers}, since
    the source terms directly depends on amplitudes and multipliers, but
    only \emph{indirectly} coupled to the amplitudes equation,
    Eq.~\eqref{eq:pted-cc-amplitudes}.
\end{enumerate}

One might exploit point 3 above to reduce the prefactor of the method to
that of an \emph{in vacuo} \acs{CC} calculation.
The simplest approach is to neglect altogether the polarization equation
and the polarization degree of freedom. This is equivalent to
differentiating the polarization-independent \acs{PTE} effective
Lagrangian:
\begin{subequations}
  \begin{align}
   \tampEq{\mu_i}(\tamp{}, \tbar{}, \p)  &=
   \denom{\mu_i}\tamp{\mu_i} + \braket{\mu_i | \overline{\Phi} | \mathrm{HF}}
    = 0 \\
   \tbarEq{\mu_i}(\tamp{}, \tbar{}, \p)
    &=
    \denom{\mu_i}\tbar{\mu_i} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} |
    \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}}
    = 0
  \end{align}
\end{subequations}
The \acs{PTE} scheme naturally preserves the scaling of the underlying \acs{CC}
method.
The \acs{PTE} scheme is advantageous since no coupling between the
\acs{CC} amplitude and multiplier equations is introduced. However, the
polarization included in the correlation treatment is the same as for
the reference determinant, an approximation which appears questionable
from the point of view of classical electrostatics.
There is, in fact, no relaxation of the reference reaction
field due to the correlated description of the electronic density.
